Dynamics of Colombo's Top: Generating Exoplanet Obliquities from Planet-Disk Interactions

### 2020-09-03 writeups/Su_disk_disp.tex
rm bad footnote per ref request

In this paper, we consider a wide range of initial planetary obliquities in the
Millholland-Batygin dissipating disk scenario, and examine how the obliquity
evolves toward the ``final'' value as the exterior disk dissipates. We provide
an analytical framework for understanding the final planetary obliquity for
arbitrary initial spin-disk misalignment angles. We also consider various
dissipation timescales, and examine both ``adiabatic'' (slow disk dissipation)
and ``non-adiabatic'' evolution. We calibrate these analytical results with
numerical calculations. \textcolor{Corr}{ On the technical side, our paper
extends previous works \citep[such as][]{henrard1982, henrard1987,
millholland_disk} in several aspects. Two of our main results are: (i) a careful
accounting of the phase space area across sepratrix to analytically describe the
rich dynamics of adiabatic evolution, and (ii) using the concept of ``partial
adiabatic resonance advection'' to fully capture the dynamics in the
non-adiabatic limit.}

@@

The spin axis $\uv{s}$ of the planet tends to precess around its orbital
(angular momentum) axis $\uv{l}$, driven by the gravitational torque from the
host star acting on the planet's rotational bulge. On the other hand, $\uv{l}$
and the disk axis $\uv{l}_{\rm d}$ precess around each other due to
gravitational interactions. We assume $S \ll L \ll L_{\rm d}$, so $\uv{l}_{\rm
d}$ is nearly constant and $\uv{l}$ experiences negligible backreaction torque
from $\uv{s}$. The equations of motion for $\uv{s}$ and $\uv{l}$ in this limit
are \textcolor{Corr}{\citep{anderson2018teeter}}
\begin{align}
    \rd{\uv{s}}{t} &= \omega_{\rm sl} \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        \equiv \alpha \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}},\label{eq:dsdt}\\
    \rd{\uv{l}}{t} &= \omega_{\rm ld}\p{\uv{l} \cdot \uv{l}_{\rm d}}
            \p{\uv{l} \times \uv{l}_{\rm d}}
        \equiv -g\p{\uv{l} \times \uv{l}_{\rm d}},
            \label{eq:dldt}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3GJ_2 M_pR_p^2 M_\star}{2a_p^3 I_p\Omega_p}
        = \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm p}}
            \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &\equiv \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
            p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
In Eq.~\eqref{eq:wsl}, $I_p = k_p M_pR_p^2$ (with $k_p$ a constant) is the
moment of inertia and $J_2=k_{\rm qp}\Omega_p^2 (R_p^3/GM_p)$ (with $k_{qp}$ a
constant) the rotation-induced (dimensionless) quadrupole of the planet [for a
body with uniform density, $k_p=0.4, k_{qp}=0.5$; for giant planets, $k_p\simeq
0.25$ and $k_{qp}\simeq 0.17$ \citep[e.g.][]{lainey2016quantification}].
\textcolor{Corr}{In other studies, $3k_{\rm qp} / 2 k_{\rm p}$ is often notated
as $k_2 / 2C$ \citep[e.g.][]{millholland_disk}}. In Eq.~\eqref{eq:wld}, $n
\equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the planet's orbital mean motion, and we
have assumed $r_d\gg a_p$ and included only the leading-order (quadrupole)
interaction between the planet and disk. We define three relative inclination
angles via
\begin{equation}
  \uv{s} \cdot \uv{l}\equiv \cos\theta,\quad
  \uv{s} \cdot \uv{l}_{\rm d}\equiv \cos\theta_{\rm sd},\quad
  \uv{l} \cdot \uv{l}_{\rm d}\equiv \cos I.
\end{equation}
In our model, $I$ is a constant.
Following standard notation \citep[e.g.][]{colombo1966,peale1969,ward2004I}, we
have defined $\alpha \equiv \omega_{\rm sl}$ and $g \equiv -\omega_{\rm ld}\cos I$.

@@

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two
trajectories originating and ending at CS4 are the only two infinite-period
orbits in the phase space. Together, these two critical trajectories are
referred to as the \emph{separatrix} and divide phase space into three zones. In
Fig.~\ref{fig:eq_1contours}, we show the separatrix, the three zones, and their
relations to the CSs. Trajectories in zone II librate about CS2 while those in
zones I and III circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{plots_diskdisp/1contours_flip.png}
    \caption{Level curves of $\mathcal{H}\p{\phi, \cos \theta}$
    [Eq.~\eqref{eq:H}] \textcolor{Corr}{for $I = 5^\circ$}, where warmer colors
    denote more positive values. The black solid line is the separatrix, which
    only exists for $\eta < \eta_{\rm c} = 0.766$. The three zones (I, II, III),
    divided by the separatrix, are labeled. The Cassini states are denoted by
    filled circles and have the same colors as in Fig.~\ref{fig:cs_locs}. The
    interior of the separatrix, shaded in grey, is formally only defined for
    $\eta < \eta_{\rm c}$, but we may identify the points in phase space that
    flow into zone II when evolved forward in time (decreasing $\eta$
    adiabatically); this is the shaded region in panel (a), bounded by the black
    dotted line.}\label{fig:eq_1contours}
\end{figure*}

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically. If we define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},
    \end{align}
\end{subequations}
\textcolor{Corr}{then the areas for $\eta < \eta_{\rm c}$ are given by
\citep{ward2004I}}
\begin{subequations}\label{se:area_ward}
    \begin{align}
        \mathcal{A}_{\rm I} &= 2\pi\p{1 - z_0} - \frac{\mathcal{A}_{\rm II}}{2},
            \label{eq:A1}\\
        \mathcal{A}_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan
            \frac{1}{\chi}, \label{eq:A2}\\
        \mathcal{A}_{\rm III} &= 2\pi\p{1 + z_0} - \frac{\mathcal{A}_{\rm
            II}}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$\mathcal{A}_{\rm j}\p{\eta > \eta_{\rm c}} = \mathcal{A}_{\rm j}(\eta_{\rm
c})$. The boundary between these extended zones is denoted by the dashed black
line in panel (a) of Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/1_areas.png}
    \caption{Plot of fractional areas of each of the zones
    $\mathcal{A}_{\rm j}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}
    \textcolor{Corr}{for $I = 5^\circ$}. The colored dotted lines correspond to
    small $\eta$ approximations used in Appendix~\ref{s:ad_approx}. The colored
    dashed lines for $\eta > \eta_{\rm c}$ are the effective values of
    $\mathcal{A}_{\rm II}, \mathcal{A}_{\rm III}$ for $\eta > \eta_{\rm c}$,
    denoting the points that would flow into either area under adiabatic
    decrease of $\eta$ from $\eta > \eta_{\rm c}$ (see the text). The vertical
    black dashed lines correspond to $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}]
    and the values of $\eta$ for which $\mathcal{A}_{\rm II}$ is maximized
    ($\eta_{\rm \max, II}$) and for which $\mathcal{A}_{\rm III}$ is minimized
    ($\eta_{\rm \min, III}$, Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_05_35.png}
    \caption{Top: The final spin obliquity $\theta_{\rm f}$ as a function of the
    initial spin-disk misalignment angle $\theta_{\rm sd, i}$ for systems
    evolving from initial $\eta_{\rm i} \gg 1$ to $\eta_{\rm f} \ll 1$,
    \textcolor{Corr}{for $I = 5^\circ$}. The blue dots are results of numerical
    calculations (Section~\ref{ss:ad_ensemble}), and the colored tracks are
    semi-analytical results (Section~\ref{ss:zone_transitions}). Bottom: The
    probabilities of different outcomes. Where a particular $\theta_{\rm sd, i}$
    corresponds to multiple tracks, the system evolves probabilistically. The
    track that a particular system evolves along in a numerical simulation can
    be measured by examining its final obliquity. The dots represent the
    inferred probabilities from measured final obliquities in our simulations,
    while the colored tracks denote the semi-analytic probability of the system
    evolving along each track. There are five regimes of $\theta_{\rm sd, i}$
    values for which different tracks are accessible. In both plots, the
    vertical dashed black lines denote semi-analytical calculations of the
    boundaries of these regimes (see Section~\ref{ss:zone_transitions}), while
    the black dotted lines represent analytical approximations valid in the
    small-$\theta_{\rm sd, i}$ limit (see
    Appendix~\ref{s:ad_approx}).}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_20_35.png}
    \caption{Same as the top panel of Fig.~\ref{fig:ad_ensemble} but for $I
    = 20^\circ$ and with fewer annotations.}\label{fig:3_ensemble_20_35}
\end{figure}

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. We adopt
convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
\textcolor{Corr}{Note that $A$ can be negative when $\rdil{\phi}{t} < 0$, unlike
the unsigned areas $\mathcal{A}_{\rm i}$ [Eqs.~\eqref{se:area_ward}] which are
positive by definition.} This definition
of $A$ has two advantages: (i) it is continuous across transitions from
circulating to librating that cross the North pole ($\cos \theta = 1$), and (ii)
the areas of the three zones are equal in absolute value to the expressions
given in Eqs.~\eqref{se:area_ward}. The path over which the integral is taken is
either a libration or circulation cycle. When $\eta_{\rm i} \gg 1$, trajectories
librate about $\uv{l}_{\rm d}$ with constant $\theta_{\rm sd}$, meaning they
enclose initial phase space area
\begin{equation}
    A_{\rm i} = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near
CS2 or CS3 librate about these equilibria, rather than $\uv{l}_{\rm d}$, and
Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice, Eq.~\eqref{eq:ai_qsd}
holds very well when defining $\theta_{\rm sd, i}$ as the angular distance to
CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

@@

When a trajectory experiences separatrix crossing, it transitions into nearby
zones probabilistically. This process is studied \textcolor{Corr}{in the
adiabatic limit} by \citet{henrard1982} and \citet{henrard1987}. Their results
may be summarized as follows: if zone $i$ is shrinking while adjacent zones $j,
k$ are expanding such that the sum of their areas is constant, the probabilities
of transition from zone $i$ to zones $j$ and $k$ are given by\footnote{
\textcolor{Corr}{Interestingly, the probabilistic nature of separatrix crossing
is not because the outcome depends on the phase $\phi$ of the system at
crossing. Instead, the transition the system takes is determined by the value of
$H$ at the start of a separatrix-crossing orbit \citep{henrard1982}. Since $H$
only changes by $\propto \epsilon$ over a circulation/libration cycle, an order
$\sim \epsilon$ change to the initial condition affects the final outcome. As
such, as $\epsilon \to 0$, the outcome effectively becomes probabilistic.}}
\begin{subequations}\label{eq:henrard_hop}
    \begin{align}
        \Pr\p{i \to j} = -\frac{\pdil{\mathcal{A}_{\rm
            j}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}, \\
        \Pr\p{i \to k} = -\frac{\pdil{\mathcal{A}_{\rm
            k}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}.
    \end{align}
\end{subequations}
Note that $\Pr \p{i \to j} + \Pr\p{i \to k} = 1$.
Eqs.~\eqref{eq:henrard_hop} can be used in conjunction with
Eqs.~\eqref{se:area_ward} to understand for what initial conditions each track
can be observed and with what probabilities.

@@

As $\epsilon$ increases (see Fig.~\ref{fig:3_ensemble_05_25}), nonadiabaticity
manifests as a larger scatter of final obliquities near the tracks predicted
from adiabatic evolution. This scatter first sets in for trajectories starting
in zone III, as these trajectories encounter the separatrix at larger $\eta$
compared to those originating in zone II\@. This means the obliquity of CS2
$\theta_2$ is smaller for these trajectories, and the adiabaticity criterion is
stricter [see Eq.~\eqref{eq:ad_constr}]. Physically, approaching the
adiabaticity criterion corresponds to the separatrix crossing process becoming
sensitive to the \emph{phase} of the libration/circulation cycle at the
crossing \textcolor{Corr}{(as noted in Section~3.2.2, the phase does not affect
separatrix crossing in the adiabatic limit)}: if the trajectory crosses the
separatrix when the obliquity is at its maximum, the final obliquity will also
be relatively larger.

As $\epsilon$ increases further (see Fig.~\ref{fig:3_ensemble_05_15}) but still
marginally satisfies the weak adiabaticity criterion [Eq.~\eqref{eq:ad_constr}],
the scatter in $\theta_{\rm f}$ continues to widen. The horizontal banded
structure of the final obliquities is a consequence of even stronger phase
sensitivity during separatrix crossing: trajectories cross the separatrix at
similar phases evolve to similar final obliquities that only depend weakly on
on $\theta_{\rm sd, i}$. \textcolor{Corr}{Finally, in
Fig.~\ref{fig:3_ensemble_05_15}, the bottom edge of the data and the III $\to$ I
track deviate very noticeably. This non-adiabatic effect is the result of the
separatrix evolving significantly within the separatrix-crossing orbit, as the
tracks computed in Section~\ref{s:ad} assume that $\eta$ is constant throughout
the separatrix-crossing orbit.}

@@

In this paper, we have studied the excitation of planetary obliquities due to
gravitational interaction with an exterior, dissipating (mass-losing)
protoplanetary disk. Obliquity excitation occurs as the system passes through a
secular resonance between spin precession and orbital (nodal) precession. This
scenario was recently studied by \citet{millholland_disk}, who focused on the
special case of small initial obliquities.  In contrast, we consider arbitrary
initial misalignment angles in this paper, motivated by the fact that planet
formation through core accretion can lead to a wide range of initial spin
orientations.  We present our result as a mapping from $\theta_{\rm sd, i}$ to
$\theta_{\rm f}$, where $\theta_{\rm sd, i}$ is the initial misalignment angle
between the planet's spin axis and the disk's orbital angular momentum axis, and
$\theta_{\rm f}$ is the final planetary obliquity. We have derived analytical
results that capture the behavior of this mapping in both the adiabatic and
nonadiabatic limits:
\begin{enumerate}
    \item In the adiabatic limit (i.e.\ the disk dissipates at a
        sufficiently slow rate), we reproduce the known result $\theta_{\rm f}
        \simeq 90^\circ$ for $\theta_{\rm sd, i} \simeq 0$. We demonstrate (via
        numerical calculation and analytical argument) the dual-valued behavior
        of $\theta_{\rm f}$ for nonzero $\theta_{\rm sd, i}$ (see
        Fig.~\ref{fig:ad_ensemble}). We \textcolor{Corr}{show for the first time
        that} both the final $\theta_{\rm f}$ values and the probabilities of
        achieving each value \textcolor{Corr}{can be understood analytically}
        via careful accounting of adiabatic invariance and separatrix crossing
        dynamics.

    \item As the disk dissipates more rapidly, the adiabatic condition
        [Eq.~\eqref{eq:ad_constr}] breaks down, we find that a broad range of
        final obliquities can be reached for a given $\theta_{\rm sd, i}$ (see
        Fig.~\ref{fig:nonad_3_ensemble}). We \textcolor{Corr}{understand this
        result via the novel concept of partial adiabatic resonance advection
        and} provide an analytical expression of the bounds on $\theta_{\rm f}$
        in Eq.~\eqref{eq:nonad_q_f_dist}.
\end{enumerate}

>>>>

In this paper, we consider a wide range of initial planetary obliquities in the
Millholland-Batygin dissipating disk scenario, and examine how the obliquity
evolves toward the ``final'' value as the exterior disk dissipates. We provide
an analytical framework for understanding the final planetary obliquity for
arbitrary initial spin-disk misalignment angles. We also consider various
dissipation timescales, and examine both ``adiabatic'' (slow disk dissipation)
and ``non-adiabatic'' evolution. We calibrate these analytical results with
numerical calculations. On the technical side, our paper
extends previous works \citep[such as][]{henrard1982, henrard1987,
millholland_disk} in several aspects. Two of our main results are: (i) a careful
accounting of the phase space area across sepratrix to analytically describe the
rich dynamics of adiabatic evolution, and (ii) using the concept of ``partial
adiabatic resonance advection'' to fully capture the dynamics in the
non-adiabatic limit.

@@

The spin axis $\uv{s}$ of the planet tends to precess around its orbital
(angular momentum) axis $\uv{l}$, driven by the gravitational torque from the
host star acting on the planet's rotational bulge. On the other hand, $\uv{l}$
and the disk axis $\uv{l}_{\rm d}$ precess around each other due to
gravitational interactions. We assume $S \ll L \ll L_{\rm d}$, so $\uv{l}_{\rm
d}$ is nearly constant and $\uv{l}$ experiences negligible backreaction torque
from $\uv{s}$. The equations of motion for $\uv{s}$ and $\uv{l}$ in this limit
are \citep{anderson2018teeter}
\begin{align}
    \rd{\uv{s}}{t} &= \omega_{\rm sl} \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}}
        \equiv \alpha \p{\uv{s} \cdot \uv{l}}
            \p{\uv{s} \times \uv{l}},\label{eq:dsdt}\\
    \rd{\uv{l}}{t} &= \omega_{\rm ld}\p{\uv{l} \cdot \uv{l}_{\rm d}}
            \p{\uv{l} \times \uv{l}_{\rm d}}
        \equiv -g\p{\uv{l} \times \uv{l}_{\rm d}},
            \label{eq:dldt}
\end{align}
where
\begin{align}
    \omega_{\rm sl} &\equiv \frac{3GJ_2 M_pR_p^2 M_\star}{2a_p^3 I_p\Omega_p}
        = \frac{3k_{\rm qp}}{2k_{\rm p}} \frac{M_\star}{m_{\rm p}}
            \p{\frac{R_{\rm p}}{a_{\rm p}}}^3 \Omega_{\rm p},\label{eq:wsl}\\
    \omega_{\rm ld} &\equiv \frac{3M_{\rm d}}{4M_\star}\p{\frac{a_{\rm
            p}}{r_{\rm d}}}^3 n .\label{eq:wld}
\end{align}
In Eq.~\eqref{eq:wsl}, $I_p = k_p M_pR_p^2$ (with $k_p$ a constant) is the
moment of inertia and $J_2=k_{\rm qp}\Omega_p^2 (R_p^3/GM_p)$ (with $k_{qp}$ a
constant) the rotation-induced (dimensionless) quadrupole of the planet [for a
body with uniform density, $k_p=0.4, k_{qp}=0.5$; for giant planets, $k_p\simeq
0.25$ and $k_{qp}\simeq 0.17$ \citep[e.g.][]{lainey2016quantification}].
In other studies, $3k_{\rm qp} / 2 k_{\rm p}$ is often notated
as $k_2 / 2C$ \citep[e.g.][]{millholland_disk}. In Eq.~\eqref{eq:wld}, $n
\equiv \sqrt{GM_\star/a_{\rm p}^3}$ is the planet's orbital mean motion, and we
have assumed $r_d\gg a_p$ and included only the leading-order (quadrupole)
interaction between the planet and disk. We define three relative inclination
angles via
\begin{equation}
  \uv{s} \cdot \uv{l}\equiv \cos\theta,\quad
  \uv{s} \cdot \uv{l}_{\rm d}\equiv \cos\theta_{\rm sd},\quad
  \uv{l} \cdot \uv{l}_{\rm d}\equiv \cos I.
\end{equation}
In our model, $I$ is a constant.
Following standard notation \citep[e.g.][]{colombo1966,peale1969,ward2004I}, we
have defined $\alpha \equiv \omega_{\rm sl}$ and $g \equiv -\omega_{\rm ld}\cos I$.

@@

When $\eta < \eta_{\rm c}$, CS4 exists and is a saddle point. The two
trajectories originating and ending at CS4 are the only two infinite-period
orbits in the phase space. Together, these two critical trajectories are
referred to as the \emph{separatrix} and divide phase space into three zones. In
Fig.~\ref{fig:eq_1contours}, we show the separatrix, the three zones, and their
relations to the CSs. Trajectories in zone II librate about CS2 while those in
zones I and III circulate.
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{plots_diskdisp/1contours_flip.png}
    \caption{Level curves of $\mathcal{H}\p{\phi, \cos \theta}$
    [Eq.~\eqref{eq:H}] for $I = 5^\circ$, where warmer colors
    denote more positive values. The black solid line is the separatrix, which
    only exists for $\eta < \eta_{\rm c} = 0.766$. The three zones (I, II, III),
    divided by the separatrix, are labeled. The Cassini states are denoted by
    filled circles and have the same colors as in Fig.~\ref{fig:cs_locs}. The
    interior of the separatrix, shaded in grey, is formally only defined for
    $\eta < \eta_{\rm c}$, but we may identify the points in phase space that
    flow into zone II when evolved forward in time (decreasing $\eta$
    adiabatically); this is the shaded region in panel (a), bounded by the black
    dotted line.}\label{fig:eq_1contours}
\end{figure*}

Since $\p{\phi, \cos \theta}$ are canonically conjugate, the integral $\oint
\cos \theta\;\mathrm{d}\phi$ along a trajectory is an adiabatic invariant (see
Section~\ref{s:ad}). The unsigned areas $\p{\abs{\int \cos \theta
\;\mathrm{d}\phi}}$ of the three zones (as defined in
Fig.~\ref{fig:eq_1contours}) can be computed analytically. If we define
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},
    \end{align}
\end{subequations}
then the areas for $\eta < \eta_{\rm c}$ are given by
\citep{ward2004I}
\begin{subequations}\label{se:area_ward}
    \begin{align}
        \mathcal{A}_{\rm I} &= 2\pi\p{1 - z_0} - \frac{\mathcal{A}_{\rm II}}{2},
            \label{eq:A1}\\
        \mathcal{A}_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan
            \frac{1}{\chi}, \label{eq:A2}\\
        \mathcal{A}_{\rm III} &= 2\pi\p{1 + z_0} - \frac{\mathcal{A}_{\rm
            II}}{2}\label{eq:A3}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. While the
zones are not formally defined for $\eta > \eta_{\rm c}$ since the separatrix
disappears, a natural extension exists: evolve an initial phase space point $p$
under adiabatic decrease of $\eta$ until the separatrix appears at $\eta =
\eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm c}$. Since
phase space area is conserved under adiabatic evolution, this extension implies
$\mathcal{A}_{\rm j}\p{\eta > \eta_{\rm c}} = \mathcal{A}_{\rm j}(\eta_{\rm
c})$. The boundary between these extended zones is denoted by the dashed black
line in panel (a) of Fig.~\ref{fig:eq_1contours}, where no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/1_areas.png}
    \caption{Plot of fractional areas of each of the zones
    $\mathcal{A}_{\rm j}(\eta) / 4\pi$ as given by Eqs.~\eqref{se:area_ward}
    for $I = 5^\circ$. The colored dotted lines correspond to
    small $\eta$ approximations used in Appendix~\ref{s:ad_approx}. The colored
    dashed lines for $\eta > \eta_{\rm c}$ are the effective values of
    $\mathcal{A}_{\rm II}, \mathcal{A}_{\rm III}$ for $\eta > \eta_{\rm c}$,
    denoting the points that would flow into either area under adiabatic
    decrease of $\eta$ from $\eta > \eta_{\rm c}$ (see the text). The vertical
    black dashed lines correspond to $\eta = \eta_{\rm c}$ [Eq.\eqref{eq:etac}]
    and the values of $\eta$ for which $\mathcal{A}_{\rm II}$ is maximized
    ($\eta_{\rm \max, II}$) and for which $\mathcal{A}_{\rm III}$ is minimized
    ($\eta_{\rm \min, III}$, Eq.~\eqref{eq:eta_minIII}).}\label{fig:eq_areas}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_05_35.png}
    \caption{Top: The final spin obliquity $\theta_{\rm f}$ as a function of the
    initial spin-disk misalignment angle $\theta_{\rm sd, i}$ for systems
    evolving from initial $\eta_{\rm i} \gg 1$ to $\eta_{\rm f} \ll 1$,
    for $I = 5^\circ$. The blue dots are results of numerical
    calculations (Section~\ref{ss:ad_ensemble}), and the colored tracks are
    semi-analytical results (Section~\ref{ss:zone_transitions}). Bottom: The
    probabilities of different outcomes. Where a particular $\theta_{\rm sd, i}$
    corresponds to multiple tracks, the system evolves probabilistically. The
    track that a particular system evolves along in a numerical simulation can
    be measured by examining its final obliquity. The dots represent the
    inferred probabilities from measured final obliquities in our simulations,
    while the colored tracks denote the semi-analytic probability of the system
    evolving along each track. There are five regimes of $\theta_{\rm sd, i}$
    values for which different tracks are accessible. In both plots, the
    vertical dashed black lines denote semi-analytical calculations of the
    boundaries of these regimes (see Section~\ref{ss:zone_transitions}), while
    the black dotted lines represent analytical approximations valid in the
    small-$\theta_{\rm sd, i}$ limit (see
    Appendix~\ref{s:ad_approx}).}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.47\textwidth]{plots_diskdisp/3_ensemble_20_35.png}
    \caption{Same as the top panel of Fig.~\ref{fig:ad_ensemble} but for $I
    = 20^\circ$ and with fewer annotations.}\label{fig:3_ensemble_20_35}
\end{figure}

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. We adopt
convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
Note that $A$ can be negative when $\rdil{\phi}{t} < 0$, unlike
the unsigned areas $\mathcal{A}_{\rm i}$ [Eqs.~\eqref{se:area_ward}] which are
positive by definition. This definition
of $A$ has two advantages: (i) it is continuous across transitions from
circulating to librating that cross the North pole ($\cos \theta = 1$), and (ii)
the areas of the three zones are equal in absolute value to the expressions
given in Eqs.~\eqref{se:area_ward}. The path over which the integral is taken is
either a libration or circulation cycle. When $\eta_{\rm i} \gg 1$, trajectories
librate about $\uv{l}_{\rm d}$ with constant $\theta_{\rm sd}$, meaning they
enclose initial phase space area
\begin{equation}
    A_{\rm i} = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near
CS2 or CS3 librate about these equilibria, rather than $\uv{l}_{\rm d}$, and
Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice, Eq.~\eqref{eq:ai_qsd}
holds very well when defining $\theta_{\rm sd, i}$ as the angular distance to
CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

@@

When a trajectory experiences separatrix crossing, it transitions into nearby
zones probabilistically. This process is studied in the
adiabatic limit by \citet{henrard1982} and \citet{henrard1987}. Their results
may be summarized as follows: if zone $i$ is shrinking while adjacent zones $j,
k$ are expanding such that the sum of their areas is constant, the probabilities
of transition from zone $i$ to zones $j$ and $k$ are given by
\begin{subequations}\label{eq:henrard_hop}
    \begin{align}
        \Pr\p{i \to j} = -\frac{\pdil{\mathcal{A}_{\rm
            j}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}, \\
        \Pr\p{i \to k} = -\frac{\pdil{\mathcal{A}_{\rm
            k}}{\eta}}{\pdil{\mathcal{A}_{\rm i}}{\eta}}.
    \end{align}
\end{subequations}
Note that $\Pr \p{i \to j} + \Pr\p{i \to k} = 1$.
Eqs.~\eqref{eq:henrard_hop} can be used in conjunction with
Eqs.~\eqref{se:area_ward} to understand for what initial conditions each track
can be observed and with what probabilities.

@@

As $\epsilon$ increases (see Fig.~\ref{fig:3_ensemble_05_25}), nonadiabaticity
manifests as a larger scatter of final obliquities near the tracks predicted
from adiabatic evolution. This scatter first sets in for trajectories starting
in zone III, as these trajectories encounter the separatrix at larger $\eta$
compared to those originating in zone II\@. This means the obliquity of CS2
$\theta_2$ is smaller for these trajectories, and the adiabaticity criterion is
stricter [see Eq.~\eqref{eq:ad_constr}]. Physically, approaching the
adiabaticity criterion corresponds to the separatrix crossing process becoming
sensitive to the \emph{phase} of the libration/circulation cycle at the
crossing: if the trajectory crosses the separatrix when the obliquity is at its
maximum, the final obliquity will also be relatively larger.

As $\epsilon$ increases further (see Fig.~\ref{fig:3_ensemble_05_15}) but still
marginally satisfies the weak adiabaticity criterion [Eq.~\eqref{eq:ad_constr}],
the scatter in $\theta_{\rm f}$ continues to widen. The horizontal banded
structure of the final obliquities is a consequence of even stronger phase
sensitivity during separatrix crossing: trajectories cross the separatrix at
similar phases evolve to similar final obliquities that only depend weakly on
on $\theta_{\rm sd, i}$. Finally, in
Fig.~\ref{fig:3_ensemble_05_15}, the bottom edge of the data and the III $\to$ I
track deviate very noticeably. This non-adiabatic effect is the result of the
separatrix evolving significantly within the separatrix-crossing orbit, as the
tracks computed in Section~\ref{s:ad} assume that $\eta$ is constant throughout
the separatrix-crossing orbit.

@@

In this paper, we have studied the excitation of planetary obliquities due to
gravitational interaction with an exterior, dissipating (mass-losing)
protoplanetary disk. Obliquity excitation occurs as the system passes through a
secular resonance between spin precession and orbital (nodal) precession. This
scenario was recently studied by \citet{millholland_disk}, who focused on the
special case of small initial obliquities.  In contrast, we consider arbitrary
initial misalignment angles in this paper, motivated by the fact that planet
formation through core accretion can lead to a wide range of initial spin
orientations.  We present our result as a mapping from $\theta_{\rm sd, i}$ to
$\theta_{\rm f}$, where $\theta_{\rm sd, i}$ is the initial misalignment angle
between the planet's spin axis and the disk's orbital angular momentum axis, and
$\theta_{\rm f}$ is the final planetary obliquity. We have derived analytical
results that capture the behavior of this mapping in both the adiabatic and
nonadiabatic limits:
\begin{enumerate}
    \item In the adiabatic limit (i.e.\ the disk dissipates at a
        sufficiently slow rate), we reproduce the known result $\theta_{\rm f}
        \simeq 90^\circ$ for $\theta_{\rm sd, i} \simeq 0$. We demonstrate (via
        numerical calculation and analytical argument) the dual-valued behavior
        of $\theta_{\rm f}$ for nonzero $\theta_{\rm sd, i}$ (see
        Fig.~\ref{fig:ad_ensemble}). We show for the first time
        that both the final $\theta_{\rm f}$ values and the probabilities of
        achieving each value can be understood analytically
        via careful accounting of adiabatic invariance and separatrix crossing
        dynamics.

    \item As the disk dissipates more rapidly, the adiabatic condition
        [Eq.~\eqref{eq:ad_constr}] breaks down, we find that a broad range of
        final obliquities can be reached for a given $\theta_{\rm sd, i}$ (see
        Fig.~\ref{fig:nonad_3_ensemble}). We understand this
        result via the novel concept of partial adiabatic resonance advection
        and provide an analytical expression of the bounds on $\theta_{\rm f}$
        in Eq.~\eqref{eq:nonad_q_f_dist}.
\end{enumerate}
